Temporal ordering and registration of imaging data in cross--sectional studies of development

### 2014-04-01 image_analysis_paper/pnas_draft.tex
Added stas comments to paper

\begin{abstract} 
In studies of development, researchers are often presented with cross-sectional data, where each data point is a sample from a population fixed at a slightly different developmental time. 
%
The goal is then to temporally order the data to reconstruct the developmental dynamics. 
%
If each data point is a two-dimensional image, the images must first be aligned or registered before they can be temporally ordered. 
%
When such data sets are large, and/or if the developmental changes are subtle, these tasks can be difficult to do by hand. 
%
We present an automatic way to register and temporally order cross--sectional data sets of images. We use relatively new mathematical techniques (angular synchronization for image registration, and diffusion maps for temporal ordering) which are applicable to a wide variety of data sets and require little {\em a priori} knowledge of the image features or the developmental dynamics. 
%
We demonstrate the utility of our methods using a collection of images from a study of {\em Drosophila} embryogenesis.
\end{abstract}

@@

\begin{figure}[t]
\includegraphics[width=8.4cm]{fig1}
\caption{Schematic illustration of reconstructing dynamics from cross-sectional data. (a) Fish, each in a different orientation and a different stage of development. (b) Fish, now registered and temporally ordered. For this example, the registration and ordering is easy to do by hand because the data set is small and the developmental changes are easy to recognize.} 
\label{fig:fish}
\customlabel{subfig:fish_unordered}{\ref{fig:fish}a}
\customlabel{subfig:fish_ordered}{\ref{fig:fish}b}
\end{figure}

The fruit fly embryo is shaped like a rice grain, approximately $1/2$~mm long. 
%
By end of the $2^{nd}$ hour of development, sequential divisions of a fertilized egg generate a system where $\sim 6,000$ nuclei are uniformly arranged in under the common plasma membrane. 
%
During the $3^{rd}$ hour, the embryo undergoes the process of cellularization, whereby membranes that enclose nuclei into different cells grow in length (Figure~\ref{subfig:membranes}). 
%
Concurrently with these morphological changes, concentration profiles of multiple regulatory molecules form inside the embryo, subdividing it into regions that give rise to future tissues and organs. 
%
Images in Figure~\ref{subfig:images_unordered} show optical sections through the short axis of the embryos. 
%
The embryos were fixed and stained with chemicals that visualize nuclei (shown in gray) as well as the spatial distributions of two different proteins (shown in green and red). 
%
Each of the images reveals a uniform arrangement of nuclei and nonuniform distribution of the two proteins, which are involved in patterning of the dorsoventral (back-to-belly) axis of the embryo.

\begin{figure}[t]
\includegraphics[width=8.4cm]{fig2}
\caption{(a) Fluorescent images of {\em Drosophila} embryos. Each embryo has been stained from Dorsal (green) and dpERK (red). (b) Selected image from data set. (c) Profile of Dorsal (green) and dpERK (red) around the circumfrence of the embryo in (b).}
\label{fig:images_unordered}
\customlabel{subfig:images_unordered}{\ref{fig:images_unordered}a}
\customlabel{subfig:one_image}{\ref{fig:images_unordered}b}
\customlabel{subfig:image_profiles}{\ref{fig:images_unordered}c}
\end{figure}

Each image in Figure~\ref{subfig:images_unordered} comes from a different embryo, captured at a different orientation and at a different and unknown point in time. 
%
The goal is to align these images in a globally consistent way and order them in a way that is most representative of the underlying developmental trajectory. 
%
Below we show that both of these tasks can be fully automated.
%
We will use angular synchronization \cite{singer2011angular}, a registration algorithm that is robust to noise, to align the images in a consistent way.
%
We will use diffusion maps \cite{coifman2005geometric}, a broadly applicable technique for the discovery of low-dimensional structures in high-dimensional data sets, to temporally order our data.
%
We will also show how both steps can be combined using vector diffusion maps \cite{singer2012vector}.
%
The paper is organized as follows. 
%
First, we use a synthetic dataset, where the correct orientation and ordering of data points is known, to formalize the tasks of image registration and temporal ordering. 
%
Then, we show that both of these tasks can be effectively combined in a single step. 
%
Following this, we return to the original set of two-dimensional images and demonstrate that our approach can readily align them and order them correctly in time. 
%
We conclude by discussing the general requirements that must be satisfied for our approach to work and compare it to other approaches and ideas in cross-sectional studies of biological dynamics.

@@

We would like to temporally order images, such as the images shown in Figure~\ref{subfig:images_unordered}.
%
We will use diffusion maps, a dimensionality reduction algorithm, to automatically order our data.
%
However, this algorithm requires distances between data points. 
%
Because each image is in a different orientation, we need to first align or {\em register} the images before computing distances.
%
Once all the images have been aligned, we can compute distances and use diffusion maps to temporally order the data.

We will first demonstrate our techniques for registration and temporal ordering using a synthetic data set with relatively simple dynamics.
%
We construct concentration profiles defined on a ring, and we assume that each ring is in an arbitrary orientation; an example is shown in Figure~\ref{subfig:1d_example}.
%
Each concentration profile is a Gaussian (shown in Figure~\ref{subfig:1d_intensity}), and the Gaussians increase in intensity as  a function of time.
%
We discritize the profiles into 100 points, so our data will be 100-dimensional vectors. 
%
Rotation of the ring in Figure~\ref{subfig:1d_example} corresponds to shifting (with periodic boundary conditions) the one-dimensional concentration profile shown below the ring. 
%
Figure \ref{subfig:1d_unaligned_unordered} shows all the concentration profiles in our data set. 
%
The concentration profiles have been stacked in an array, such that each row corresponds to one data point.
%
Because the profiles are unregistered and unordered, the underlying dynamics (a Gaussian growing in time) are not readily apparent.

\begin{figure}[t]
%\includegraphics{figures_0329_mac_4}
\includegraphics[width=8.4cm]{fig4}
\caption{Illustration of methodologies using one-dimensional profiles. (a) One-dimensional concentration profile on a ring (top), and the corresponding profile on a line (bottom). Rotation of the ring corresponds to shifting the line left and right (with periodic boundary conditions). (b) One-dimensional profile correspondsing to the profile in (a). This profile is similar to the profile in Figure~\ref{fig:images_unordered}(c).  (c) Many concentration profiles, each of the form described in (a). Each row in the array corresponds to one image. The overall dynamics are a Gaussian bump growing in time, but this is not readily apparent from the data because the profiles are unregistered and unordered. (d) The concentration profiles in (b), now registered using angular synchronization. (e) The concentration profiles in (c), now temporally ordered using diffusion maps. The profiles are sorted according to the value of the first non-trivial DMAPS component, $\phi_2$. (f) The concentration profiles in (b), now simultaneously registered and temporally ordered using vector diffusion maps. The profiles are sorted according to the value of the first non-trivial VDM component, $\psi_{1, 3}$.}
\label{fig:1d_demo}
\customlabel{subfig:1d_example}{\ref{fig:1d_demo}a}
\customlabel{subfig:1d_intensity}{\ref{fig:1d_demo}b}
\customlabel{subfig:1d_unaligned_unordered}{\ref{fig:1d_demo}c}
\customlabel{subfig:1d_aligned_unordered}{\ref{fig:1d_demo}d}
\customlabel{subfig:1d_aligned_ordered}{\ref{fig:1d_demo}e}
\customlabel{subfig:1d_aligned_ordered_vdm}{\ref{fig:1d_demo}f}
\end{figure}

@@

Our first task is to register the set of images or profiles.
%
%``Template-based'' registration \cite{ahuja2007template} is a common technique, where one selects a specific image or {\em template}, and then translates and rotates each image in the data set until it is optimally aligned with this template (optimal for us means minimizing the norm between the difference in pixel intensities).
%
%However, this method can result in many errors if the images are noisy, and can be problematic if a ``good'' template image is not known {\em a priori}. 
%
We will use a relatively new technique called angular synchronization\cite{singer2011angular}, which is robust to noise and requires no {\em a priori} knowledge of image features.
%
Angular synchronization first computes the pairwise alignments %(rotations for our one-dimensional profiles in Figure \ref{fig:1d_demo}c, or translations and rotations for the two-dimensional images in Figure \ref{fig:images_unordered}) 
for every possible pair of data points. %, effectively using each image as a template for every other image.
%
An illustration is shown in Figure~\ref{subfig:synch1};
each data point is a vector, and $\theta_{ij}$ denotes the angle of rotation needed to align vectors $i$ and $j$ (these are the pairwise alignments).
%
From this large collection of pairwise alignments, angular synchronization then rotates each data point, such that the individual alignments are most consistent (in some metric) with the pairwise alignments.
% (i.e., we would like the difference between the optimal rotations for two different concentration profiles to be as close as possible to the pairwise rotation needed to align the two profiles).
%
Figure~\ref{subfig:synch2} shows each vector from Figure~\ref{subfig:synch1}, now rotated so that the entire set is aligned.
%-- these are the angles we would like to recover from the pairwise information. 
%
Note that the differences between the rotation angles are consistent with the pairwise angles needed to rotate one vector to another, i.e., $\theta_i - \theta_j \approx \theta_{ij}$.
%
To compute these alignments, we formulate the task as an optimization problem, and then relax it to an eigenproblem.
%
The top eigenvector(s) of a matrix (which contains information about the pairwise alignments) give the alignments for each data point.
%
%Because the method considers {\em all} pairwise alignments, angular synchronization is often robust to noise.

\begin{figure}[t]
\includegraphics[width=8.4cm]{fig3}
\caption{Illustration of diffusion maps and angular synchronization (a) Data points (in blue) which lie on a one-dimensional nonlinear curve in two dimensions. Each pair of data points is connected by an edge, and the edge intensity is related to the distance between the points (close data points are connected by dark edges). (b) The data in (a), colored by the first (non-trivial) eigenvector from diffusion maps, $\phi_2$. The color parameterizes the one-dimensional curve. (c) Set of vectors, each in a different orientation. The angles to align one vector to another are indicated. (d) The vectors from (c), each rotated so that the set of vectors is aligned. Note that the angles are consistent with the pairwise alignments in (c). }
\label{fig:schematics}
\customlabel{subfig:dmaps_edges}{\ref{fig:schematics}a}
\customlabel{subfig:dmaps_color}{\ref{fig:schematics}b}
\customlabel{subfig:synch1}{\ref{fig:schematics}c}
\customlabel{subfig:synch2}{\ref{fig:schematics}d}
\end{figure}

Figure \ref{subfig:1d_aligned_unordered} shows the concentration profiles from Figure~\ref{subfig:1d_unaligned_unordered}, now aligned using angular synchronization.
%
The peaks are now aligned, even though we input no information about the existence or location of these peaks into our algorithm.
%
We would like to note that, because angular synchronization only uses pairwise rotation information, the optimal rotations are recovered up to a global rotation.

@@

After registering the images or concentration profiles, we want to order them in time to reconstuct the developmental dynamics. 
%
We will use diffusion maps (DMAPS) \cite{coifman2005geometric} to temporally order our data.
%
In general, diffusion maps is a dimensionality reduction technique, which aims to uncover low-dimensional, nonlinear structure in a high-dimensional data set. 
%
Figure~\ref{subfig:dmaps_edges} shows an example of a two-dimensional data set (the ``high-dimensional data'') which lies on a one-dimensional nonlinear curve (the low-dimensional structure).
%
Figure~\ref{subfig:dmaps_color} shows the same data, now colored by the first diffusion maps coordinate.
%
Clearly, the color parameterizes the one-dimensional curve that our eye can visually detect.
%
We assume that our imaging data also falls on a one-dimensional, nonlinear curve, and that this curve is parameterized by time (e.g., in Figure~\ref{subfig:dmaps_color}, we assume that the data points are ordered from youngest to oldest along the curve). 
%
Therefore, ordering our data along this curve will effectively order our data in time.
%
Because our data will be very high dimensional, we cannot do this by eye, and instead require an algorithm to automatically extract this one-dimensional structure.

To uncover a parameterization of this curve, we assume that data points which are close in the high-dimensional space should also be close on the curve (we would like images which look similar to be close together in our developmental trajectory).
%
%We therefore want to find a parameterization of the points, such that the points which are close in the original, high-dimensional space are {\em as close as possible} in the parameterization.
%
As an illustration of how local neighbors are important for parameterizing a curve, the edges between the pairs of data points in Figure~\ref{subfig:dmaps_edges} are colored according to the distance between the data points, so that close data points have a dark edge between them.
%
Visually, the dark edges are meaningful and delineate the one-dimensional curve.
%
Furthermore, the light edges are not informative and do not depict any meaningful structure. 
%
We want points which are close in the high-dimensional space, i.e., connected by dark edges, to be close in our low-dimensional parameterization.
%
This problem can be formulated as an optimization problem \cite{Belkin2003}, and the solution is given by the top (non-trivial) eigenvector of a matrix which contains information about pairwise distances between the data points. 
%
In general, this procedure is called diffusion maps \cite{coifman2005geometric}.%; the formulation extends from curves to manifolds (i.e., we can search for an $n$-dimensional, rather than a one-dimensional, parameterization of the data, such that data points which are close in the high-dimensional space are also close in the $n$-dimensional parameterization).

Given our set of registered signals, shown in Figure \ref{subfig:1d_aligned_unordered}, we can use diffusion maps to order them in time.
%
We use the Euclidean distance between the aligned concentration profiles (the rows of Figure \ref{subfig:1d_aligned_unordered}) as our pairwise distances between data points.
%
Figure \ref{subfig:1d_aligned_ordered} shows the data from Figure \ref{subfig:1d_aligned_unordered}, now sorted by the value of the first (non-trivial) eigenvector, $\phi_2$. 
%
As expected, the intensity of the peak grows as a function of time. 
%
Although diffusion maps can order the data, it cannot tell us about the speed of this trajectory.
%
Furthermore, the direction of the trajectory is also unknown; determining which end of the trajectory corresponds to the beginning of development is a task that needs to be done by hand.

@@

We are often interested in both registering and ordering our data.
%
%Our proposed alignment methodology, angular synchronization, utilizes the eigendecompostion of the matrix $H \in \mathbb{R}^{md \times md}$, and our proposed ordering algorithm, diffusion maps, utilizes the eigendecomposition of the matrix $A \in \mathbb{R}^{m \times m}$.
%
We can combine angular synchronization (for registration) and diffusion maps (for temporal ordering) into one eigencomputation that allows us to {\em simultaneously} recover the optimal alignments and the temporal ordering.
%
This technique is called vector diffusion maps \cite{singer2012vector}.
%
It both reduces the number of required eigencomputations, and is much more robust to noise, as mistakes in alignment are also considered when ordering the data.
%
Figure \ref{subfig:1d_aligned_ordered_vdm} shows the result of simultaneously registering and ordering the data in Figure \ref{subfig:1d_unaligned_unordered} using vector diffusion maps. 
%
The data has been ordered by $\psi_{1, 3}$, the first (non-trivial) vector diffusion maps coordinate.

@@

In these {\em Drosophila} experiments, both registration and temporal ordering can be done by hand using prior knowledge about the system. 
%
It is known that the peak in the distribution of Dorsal (visualized by the green channel) specifies the ventralmost position in the embryo, and that the nuclei (shown in gray) delineate the circumfrence of the embryo. 
%
Based on this, images can be translated to make the centers of the embryos coincide and rotated to place the peak of the green channel at the same angular position. 
%
At the same time, temporal ordering can be done based on the monotonic progress of cellularization (Figure~\ref{subfig:membranes}). 
%
We can evaluate the quality our automatic orderings by comparing with the orderings obtained from cellularization.
%
The rank correlation coefficient between the two orderings is 0.9149, indicating that our automatic ordering is consistent with the ordering obtained from cellularization.
%
We can also visually compare the alignments and orderings by comparing the arrays of one-dimensional profiles, similar to the format in Figure~\ref{fig:1d_demo};
these results are shown in Figures~\ref{subfig:1d_membranes}~and~\ref{subfig:1d_vdm}.
%
The general trends are consistent:
the green peaks are aligned, and the red peaks grow as a function of time. 
%
However, Figure~\ref{subfig:1d_vdm}, which was registered and ordered using our mathematical algorithms, shows a better clustering of the outer peaks in the top corners of the image.
%
%These peaks are more scrambled in Figure~\ref{fig:membrane_compare}b, which was analyzed using cellularization.
%
These peaks are known to arise later in development, and so it is reassuring that they are at the end of the trajectory.
%
Furthermore, the membranes stop growing towards the end of this trajectory, making the ordering of the late signals in Figure~\ref{subfig:1d_membranes} less accurate.

@@

\begin{figure}[t]
\includegraphics[width=8.4cm]{fig5}
\caption{The images in Figure~\ref{fig:images_unordered}, aligned and temporally ordered using vector diffusion maps. The images are sorted according to the value of the first non-trivial vector diffusion maps component, $\psi_{1, 4}$.}
\label{fig:images_ordered}
\end{figure}

\begin{figure}[t]
%\includegraphics{figures_0329_mac_6}
\includegraphics[width=8.4cm]{fig6}
\caption{(a) Images, stained for membranes (gray) and Dorsal (green) at different levels of development. (b) Membrane length as a function of time. (c) Distribution of embryo ages in the data set. (d) The one-dimensional concentration profiles, extracted, registered, and ordered by hand. The profiles are extracted using the nuclei positions in the image, registered using the location of the Dorsal peak, and ordered using cellularization. (e) The one-dimensional concentration profiles corresponding to the images in Figure~\ref{fig:images_ordered}. }
\label{fig:membrane_compare}
\customlabel{subfig:membranes}{\ref{fig:membrane_compare}a}
\customlabel{subfig:membrane_curve}{\ref{fig:membrane_compare}b}
\customlabel{subfig:membrane_dist}{\ref{fig:membrane_compare}c}
\customlabel{subfig:1d_membranes}{\ref{fig:membrane_compare}d}
\customlabel{subfig:1d_vdm}{\ref{fig:membrane_compare}e}
\end{figure}

@@

We have presented mathematical techniques for registration and temporal ordering of images arising in studies of developmental biology.
%
Although we illustrated our methods using a specific data set from a study of {\em Drosophila} embryogenesis, our framework is very general and requires little knowledge of the image features or the developmental dynamics.
%
To our knowledge, temporal ordering of {\em Drosophila} images has not been automated to this extent.
%
Recently, machine learning algorithms have been used to group {\em Drosophila} images into different developmental classes \cite{yuan2014automated}.
%
However, these methods have the disadvantages that the user must define the appropriate classes, and that hand-labeled training data is required.

>>>>

\begin{abstract} 
In studies of development, researchers are often presented with cross-sectional data, where each data point is a sample from a population fixed at a slightly different developmental time. 
%
The goal is then to temporally order the data to reconstruct the developmental dynamics. 
%
If each data point is a two-dimensional image, the images must first be aligned or registered before they can be temporally ordered. 
%
When such data sets are large, and/or if the developmental changes are subtle, these tasks can be difficult to do by hand. 
%
We present an automatic way to register and temporally order cross--sectional data sets of images. 
%
We use mathematical techniques (angular synchronization for image registration, and diffusion maps for temporal ordering) which are applicable to a wide variety of data sets and require little {\em a priori} knowledge of the image features or the developmental dynamics. 
%
We demonstrate the utility of our methods using a collection of images from a study of {\em Drosophila} embryogenesis.
\end{abstract}

@@

\begin{figure}[t]
\includegraphics[width=8.4cm]{fig1}
\caption{Schematic illustration of reconstructing dynamics from cross-sectional data. (a) Fish, each in a different orientation and a different stage of development. (b) Fish, now registered and temporally ordered. For this example, the registration and ordering is easy to do by hand because the data set is small and the developmental changes are easy to recognize.} 
%\label{fig:fish}
\customlabel{fig:fish}{1}
\customlabel{subfig:fish_unordered}{\ref{fig:fish}a}
\customlabel{subfig:fish_ordered}{\ref{fig:fish}b}
\end{figure}

The fruit fly embryo is shaped like a rice grain, approximately $1/2$~mm long. 
%
By end of the $2^{nd}$ hour of development, sequential divisions of a fertilized egg generate a system where $\sim 6,000$ nuclei are uniformly arranged in under the common plasma membrane. 
%
During the $3^{rd}$ hour, the embryo undergoes the process of cellularization, whereby membranes that enclose nuclei into different cells grow in length (Figure~\ref{subfig:membranes}). 
%
Concurrently with these morphological changes, concentration profiles of multiple regulatory molecules form inside the embryo, subdividing it into regions that give rise to future tissues and organs \cite{lim2013kinetics}. 
%
Images in Figure~\ref{subfig:images_unordered} show optical sections through the short axis of the embryos. 
%
The embryos were fixed and stained with chemicals that visualize nuclei (shown in gray) as well as the spatial distributions of two different proteins (shown in green and red) \cite{chung2010microfluidic}. 
%
Each of the images reveals a uniform arrangement of nuclei and nonuniform distribution of the two proteins, which are involved in patterning of the dorsoventral (back-to-belly) axis of the embryo.

\begin{figure}[t]
\includegraphics[width=8.4cm]{fig2}
\caption{(a) Fluorescent images of {\em Drosophila} embryos. Each embryo has been stained from Dorsal (green) and dpERK (red). (b) Selected image from data set. (c) Profile of Dorsal (green) and dpERK (red) around the circumfrence of the embryo in (b).}
%\label{fig:images_unordered}
\customlabel{fig:images_unordered}{2}
\customlabel{subfig:images_unordered}{\ref{fig:images_unordered}a}
\customlabel{subfig:one_image}{\ref{fig:images_unordered}b}
\customlabel{subfig:image_profiles}{\ref{fig:images_unordered}c}
\end{figure}

Each image in Figure~\ref{subfig:images_unordered} comes from a different embryo, captured at a different orientation and at a different and unknown point in time. 
%
The goal is to align these images in a globally consistent way and order them in a way that is most representative of the underlying developmental trajectory. 
%
Below we show that both of these tasks can be fully automated.
%
We will use angular synchronization \cite{singer2011angular}, a registration algorithm that is robust to noise, to align the images in a consistent way.
%
We will use diffusion maps \cite{coifman2005geometric}, a broadly applicable technique for the discovery of low-dimensional structures in high-dimensional data sets, to temporally order our data.
%
We will also show how both steps can be combined using vector diffusion maps \cite{singer2012vector}.
%
%The paper is organized as follows. 
%%
%First, we use a synthetic dataset, where the correct orientation and ordering of data points is known, to formalize the tasks of image registration and temporal ordering. 
%%
%Then, we show that both of these tasks can be effectively combined in a single step. 
%%
%Following this, we return to the original set of two-dimensional images and demonstrate that our approach can readily align them and order them correctly in time. 
%%
%We conclude by discussing the general requirements that must be satisfied for our approach to work and compare it to other approaches and ideas in cross-sectional studies of biological dynamics.

@@

%We would like to temporally order images, such as the images shown in Figure~\ref{subfig:images_unordered}.
%%
%We will use diffusion maps, a dimensionality reduction algorithm, to automatically order our data.
%%
%However, this algorithm requires distances between data points. 
%%
%Because each image is in a different orientation, we need to first align or {\em register} the images before computing distances.
%%
%Once all the images have been aligned, we can compute distances and use diffusion maps to temporally order the data.

We will first demonstrate our techniques for registration and temporal ordering using a synthetic data set with relatively simple dynamics.
%
We construct concentration profiles defined on a ring, and we assume that each ring is in an arbitrary orientation; an example is shown in Figure~\ref{subfig:1d_example}.
%
Each concentration profile is a noisy Gaussian (shown in Figure~\ref{subfig:1d_intensity}), and the Gaussians increase in intensity as  a function of time.
%
We discritize the profiles into 100 points, so our data will be 100-dimensional vectors. 
%
Rotation of the ring in Figure~\ref{subfig:1d_example} corresponds to shifting (with periodic boundary conditions) the one-dimensional concentration profile shown below the ring. 
%
Figure \ref{subfig:1d_unaligned_unordered} shows all the concentration profiles in our data set. 
%
The concentration profiles have been stacked in an array, such that each row corresponds to one data point.
%
Because the profiles are unregistered and unordered, the underlying dynamics (a Gaussian growing in time) are not readily apparent.

\begin{figure}[t]
%\includegraphics{figures_0329_mac_4}
\includegraphics[width=8.4cm]{fig4}
\caption{Illustration of methodologies using one-dimensional profiles. (a) One-dimensional concentration profile on a ring (top), and the corresponding profile on a line (bottom). (b) Intensity corresponding to the profile in (a). (c) Many concentration profiles, each of the form described in (a). Each row in the array corresponds to one image. (d) The concentration profiles in (b), now registered using angular synchronization. (e) The concentration profiles in (c), now temporally ordered using diffusion maps. (f) The concentration profiles in (b), now simultaneously registered and temporally ordered using vector diffusion maps.}
%\label{fig:1d_demo}
\customlabel{fig:1d_demo}{3}
\customlabel{subfig:1d_example}{\ref{fig:1d_demo}a}
\customlabel{subfig:1d_intensity}{\ref{fig:1d_demo}b}
\customlabel{subfig:1d_unaligned_unordered}{\ref{fig:1d_demo}c}
\customlabel{subfig:1d_aligned_unordered}{\ref{fig:1d_demo}d}
\customlabel{subfig:1d_aligned_ordered}{\ref{fig:1d_demo}e}
\customlabel{subfig:1d_aligned_ordered_vdm}{\ref{fig:1d_demo}f}
\end{figure}

@@

Our first task is to register the set of images or profiles.
%
%``Template-based'' registration \cite{ahuja2007template} is a common technique, where one selects a specific image or {\em template}, and then translates and rotates each image in the data set until it is optimally aligned with this template (optimal for us means minimizing the norm between the difference in pixel intensities).
%
%However, this method can result in many errors if the images are noisy, and can be problematic if a ``good'' template image is not known {\em a priori}. 
%
We will use angular synchronization\cite{singer2011angular}, a technique which is robust to noise and requires no {\em a priori} knowledge of image features.
%
Angular synchronization first computes the pairwise alignments %(rotations for our one-dimensional profiles in Figure \ref{fig:1d_demo}c, or translations and rotations for the two-dimensional images in Figure \ref{fig:images_unordered}) 
for every possible pair of data points. %, effectively using each image as a template for every other image.
%
An illustration is shown in Figure~\ref{subfig:synch1};
each data point is a vector, and $\theta_{ij}$ denotes the angle of rotation needed to align vectors $i$ and $j$ (these are the pairwise alignments).
%
From this large collection of pairwise alignments, angular synchronization then rotates each data point, such that the individual alignments are most consistent (in some metric) with the pairwise alignments.
% (i.e., we would like the difference between the optimal rotations for two different concentration profiles to be as close as possible to the pairwise rotation needed to align the two profiles).
%
Figure~\ref{subfig:synch2} shows each vector from Figure~\ref{subfig:synch1}, now rotated so that the entire set is aligned.
%-- these are the angles we would like to recover from the pairwise information. 
%
Note that the differences between the rotation angles are consistent with the pairwise angles needed to rotate one vector to another, i.e., $\theta_i - \theta_j \approx \theta_{ij}$.
%
To compute these alignments, we formulate the task as an optimization problem, and then relax it to an eigenproblem.
%
The top eigenvector(s) of a matrix (which contains information about the pairwise alignments) give the alignments for each data point.
%
%Because the method considers {\em all} pairwise alignments, angular synchronization is often robust to noise.

\begin{figure}[t]
\includegraphics[width=8.4cm]{fig3}
\caption{Illustration of diffusion maps and angular synchronization (a) Data points (in blue) which lie on a one-dimensional nonlinear curve in two dimensions. Each pair of data points is connected by an edge, and the edge intensity is related to the distance between the points (close data points are connected by dark edges). (b) The data in (a), colored by the first (non-trivial) eigenvector from diffusion maps, $\phi_2$. The color parameterizes the one-dimensional curve. (c) Set of vectors, each in a different orientation. The angles to align one vector to another are indicated. (d) The vectors from (c), each rotated so that the set of vectors is aligned. Note that the angles are consistent with the pairwise alignments in (c). }
%\label{fig:schematics}
\customlabel{fig:schematics}{4}
\customlabel{subfig:dmaps_edges}{\ref{fig:schematics}a}
\customlabel{subfig:dmaps_color}{\ref{fig:schematics}b}
\customlabel{subfig:synch1}{\ref{fig:schematics}c}
\customlabel{subfig:synch2}{\ref{fig:schematics}d}
\end{figure}

Figure \ref{subfig:1d_aligned_unordered} shows the concentration profiles from Figure~\ref{subfig:1d_unaligned_unordered}, now aligned using angular synchronization.
%
The peaks are now aligned, even though our algorithm use no information about the existence or location of these peaks.
%
Because angular synchronization only uses pairwise rotation information, the optimal rotations are recovered up to a global rotation.

@@

After registering the images or concentration profiles, we want to order them in time to reconstuct the developmental dynamics. 
%
We will use diffusion maps (DMAPS) to temporally order our data.
%
In general, diffusion maps is a dimensionality reduction technique, which aims to uncover low-dimensional, nonlinear structure in a high-dimensional data set \cite{coifman2005geometric}. 
%
Figure~\ref{subfig:dmaps_edges} shows an example of a two-dimensional data set (the ``high-dimensional data'') which lies on a one-dimensional nonlinear curve (the low-dimensional structure).
%
Figure~\ref{subfig:dmaps_color} shows the same data, now colored by the first diffusion maps coordinate.
%
The color parameterizes the one-dimensional curve that our eye can visually detect.
%
We assume that our imaging data also falls on a one-dimensional, nonlinear curve, and that this curve is parameterized by time (e.g., in Figure~\ref{subfig:dmaps_color}, we assume that the data points are ordered from youngest to oldest along the curve). 
%
Therefore, ordering data along this curve will effectively order data in time.
%
Because our data will be very high dimensional, we cannot do this by eye, and instead require an algorithm to automatically extract this one-dimensional structure.

To uncover a parameterization of this curve, we assume that data points which are close in the high-dimensional space should also be close on the curve (we would like images which look similar to be close together in our developmental trajectory).
%
%We therefore want to find a parameterization of the points, such that the points which are close in the original, high-dimensional space are {\em as close as possible} in the parameterization.
%
To emphasize this point, edges between the pairs of data points in Figure~\ref{subfig:dmaps_edges} are colored according to the distance between the data points, so that close data points have a dark edge between them.
%
Visually, the dark edges are meaningful and delineate the one-dimensional curve.
%
Furthermore, the light edges are not informative and do not depict any meaningful structure. 
%
We want points which are close in the high-dimensional space, i.e., connected by dark edges, to be close in our low-dimensional parameterization.
%
This problem can be formulated as an optimization problem \cite{Belkin2003}, and the solution is given by the top (non-trivial) eigenvector of a matrix which contains information about pairwise distances between the data points. 
%
In general, this procedure is called diffusion maps \cite{coifman2005geometric}.%; the formulation extends from curves to manifolds (i.e., we can search for an $n$-dimensional, rather than a one-dimensional, parameterization of the data, such that data points which are close in the high-dimensional space are also close in the $n$-dimensional parameterization).

Given our set of registered signals, shown in Figure \ref{subfig:1d_aligned_unordered}, we can use diffusion maps to order them in time.
%
We use the Euclidean distance between the aligned concentration profiles (the rows of Figure \ref{subfig:1d_aligned_unordered}) as our pairwise distances between data points.
%
Figure \ref{subfig:1d_aligned_ordered} shows the data from Figure \ref{subfig:1d_aligned_unordered}, now sorted by the value of the first (non-trivial) eigenvector, $\phi_2$. 
%
As expected, the intensity of the peak grows as a function of time. 
%
%Although diffusion maps can order the data, it cannot tell us about the speed of this trajectory.
%
%Furthermore, the direction of the trajectory is also unknown; determining which end of the trajectory corresponds to the beginning of development is a task that needs to be done by hand.

@@

%We are often interested in both registering and ordering our data.
%
%Our proposed alignment methodology, angular synchronization, utilizes the eigendecompostion of the matrix $H \in \mathbb{R}^{md \times md}$, and our proposed ordering algorithm, diffusion maps, utilizes the eigendecomposition of the matrix $A \in \mathbb{R}^{m \times m}$.
%
We can combine angular synchronization (for registration) and diffusion maps (for temporal ordering) into one eigencomputation that allows us to {\em simultaneously} recover the optimal alignments and the temporal ordering.
%
This technique is called vector diffusion maps \cite{singer2012vector}.
%
It both reduces the number of required eigencomputations, and is much more robust to noise, as mistakes in alignment are also considered when ordering the data.
%
Figure \ref{subfig:1d_aligned_ordered_vdm} shows the result of simultaneously registering and ordering the data in Figure \ref{subfig:1d_unaligned_unordered} using vector diffusion maps. 
%
The data has been ordered by $\psi_{1, 3}$, the first (non-trivial) vector diffusion maps coordinate.

@@

In these experiments, we can evaluate the quality of our automatic registration and ordering using prior knowledge about the system.
%
It is known that the peak in the distribution of Dorsal (visualized by the green channel) specifies the ventralmost position in the embryo, and that the nuclei (shown in gray) delineate the circumference of the embryo. 
%
%Based on this, images can be translated to make the centers of the embryos coincide and rotated to place the peak of the green channel at the same angular position. 
%
%At the same time, 
Temporal ordering can be done based on the monotonic progress of cellularization (Figure~\ref{subfig:membranes}) \cite{figard2013plasma}. 
%
We can evaluate the quality our automatic orderings by comparing with the orderings obtained from cellularization.
%
The rank correlation coefficient between the two orderings is 0.9149, indicating that our automatic ordering is consistent with the ordering obtained from cellularization.
%
We can also visually compare the alignments and orderings by comparing the arrays of one-dimensional profiles, similar to the format in Figure~\ref{fig:1d_demo};
these results are shown in Figures~\ref{subfig:1d_membranes}~and~\ref{subfig:1d_vdm}.
%
The general trends are consistent:
the green peaks are aligned, and the red peaks grow as a function of time. 
%
However, Figure~\ref{subfig:1d_vdm}, which was automatically registered and ordered, shows a better clustering of the outer peaks in the top corners of the image.
%
%These peaks are more scrambled in Figure~\ref{fig:membrane_compare}b, which was analyzed using cellularization.
%
These peaks are known to arise later in development, and so it is reassuring that they are at the end of the trajectory.
%
%Furthermore, the membranes stop growing towards the end of this trajectory, making the ordering of the late signals in Figure~\ref{subfig:1d_membranes} less accurate.

@@

\begin{figure}[t]
\includegraphics[width=8.4cm]{fig5}
\caption{The images in Figure~\ref{subfig:images_unordered}, aligned and temporally ordered using vector diffusion maps. The images are sorted according to the value of the first non-trivial vector diffusion maps component, $\psi_{1, 4}$.}
%\label{fig:images_ordered}
\customlabel{fig:images_ordered}{5}
\end{figure}

\begin{figure}[t]
%\includegraphics{figures_0329_mac_6}
\includegraphics[width=8.4cm]{fig6}
\caption{(a) Images, stained for membranes (gray) and Dorsal (green) at different levels of development. (b) Membrane length as a function of time. (c) The one-dimensional concentration profiles, extracted, registered, and ordered by hand. The profiles are extracted using the nuclei positions in the image, registered using the location of the Dorsal peak, and ordered using cellularization. (d) The one-dimensional concentration profiles corresponding to the images in Figure~\ref{fig:images_ordered}. }
%\label{fig:membrane_compare}
\customlabel{fig:membrane_compare}{6}
\customlabel{subfig:membranes}{\ref{fig:membrane_compare}a}
\customlabel{subfig:membrane_curve}{\ref{fig:membrane_compare}b}
\customlabel{subfig:1d_membranes}{\ref{fig:membrane_compare}c}
\customlabel{subfig:1d_vdm}{\ref{fig:membrane_compare}d}
\end{figure}

@@

We have shown how dimensionality reduction techniques can greatly acceleration the registration and temporal ordering of cross-sectional imaging data.
%
Although we illustrated our methods using a specific data set from a study of {\em Drosophila} embryogenesis, our framework is very general and requires little knowledge of the image features or the developmental dynamics.
%
To our knowledge, temporal ordering of {\em Drosophila} images has not been automated to this extent.
%
Recently, machine learning algorithms have been used to group {\em Drosophila} images into different developmental classes \cite{yuan2014automated}.
%
However, these methods have the disadvantages that the user must define the appropriate classes, and that hand-labeled training data is required.
